Bayesian regression modelling facilitates quantitative modelling of cell metabolism

### 2023-10-28 paper/main.qmd
Update paper and supplementary, add main .gitignore

The Laplace approximation yielded XXX samples in XXX time. The diagnostics
indicated that our algorithm was able to find the maximum a posteriori parameter
configuration, approximate the Hessian and use these quantities to generate
approximate posterior samples. The results can be found at <https://github.com/biosustain/Methionine_model/tree/main/results>.

@fig-laplace summarises the results of comparing the Laplace approximation of our
case study’s posterior distribution with the true posterior. As can be seen
from the top left plot, the Laplace method does not provide a good approximation
to the true posterior distribution, as the marginal distribution of the
total log probability density is clearly different. This was confirmed using
the Kolmogorov-Smirnov test, which is a test to differentiate two empirical
univariate distributions. The two distributions were significantly different
with a p-value indistinguishable from zero.

The difference between the Laplace approximation output and the true posterior
distribution manifests itself not only in the parameter space, but also in
the measurement space. Figure 5 frame B compares the 5%-95% interval for
flux measurement log likelihoods in the true posterior with the Laplace
approximation; lower log likelihood values indicate that the modelled and
measured values are further away. The graph shows that the Laplace approximation
yielded significantly worse predictions than the true posterior, even for the
training data.

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants [LATEX CONSTANT]- and [LATEX CONSTANT]: see
the bottom right cell of figure 5. This comparison demonstrates that the Laplace
method is not able to capture the correct relationships between parameters’
distributions.

This result shows that MCMC, while slower than Laplace approximation, is
unfortunately preferable for posterior inference in this case. We expect that
this is typical of kinetic models of realistic metabolic networks in general, so
we recommend that Maud users use MCMC sampling if possible.

>>>>

Our overall strategy was to compare samples generated using MCMC, which we
were confident could be treated as draws from the true posterior distribution,
with approximate posterior samples generated using the Laplace method. We were
unable to generate approximate posterior samples for our main methionine cycle
case study using Laplace approximation, as the algorithm could not recover from
solver failures caused by unrealistic parameter configurations. We therefore
made a comparison for a simpler model. This input can be found at <https://github.com/biosustain/Methionine_model/tree/main/data/example_ode>.

MCMC sampling for our comparison input yielded XXX samples in XXX time; Laplace
sampling yielded XXX samples in XXX time. The diagnostics indicated that our
algorithm was able to find the maximum a posteriori parameter configuration,
approximate the Hessian and use these quantities to generate approximate
posterior samples. The results can be found at <https://github.com/biosustain/Methionine_model/tree/main/results>.

@fig-laplace summarises the results of comparing the samples generated using
each method. As can be seen from the top left plot, the Laplace method does not
provide a good approximation to the true posterior distribution in this case,
as the marginal distribution of the total log probability density is clearly
different. This impression was confirmed using the Kolmogorov-Smirnov test,
which is a test to differentiate two empirical univariate distributions. The
two distributions were significantly different with a p-value indistinguishable
from zero.

The difference between the Laplace approximation output and the true posterior
distribution manifests not only in the parameter space, but also in the
measurement space. Figure @fig-laplace frame B compares the 5%-95% interval
for flux measurement log likelihoods in the true posterior with the Laplace
approximation; lower log likelihood values indicate that the modelled and
measured values are further away. The graph shows that the Laplace approximation
yielded significantly worse predictions than the true posterior, even for the
training data.

To further explore why this is the case we compared samples from the true
posterior and the Laplace approximation for the pairwise marginal distributions
of two Michaelis-Menten constants [XXX]- and [XXX]: see the bottom right cell of
figure @fig-laplace. This comparison demonstrates that the Laplace method is not
able to capture the correct relationships between parameters’ distributions.

This result shows that MCMC, while slower than Laplace approximation, is
unfortunately preferable for posterior inference in this case. We expect that
the Laplace method will produce worse approximations the more complex the target
model. Since the approximation is already unacceptable for our simple test
model, we recommend that Maud users use MCMC sampling in preference to Laplace
approximation if possible when fitting realistic Bayesian kinetic models.

### 2023-10-28 paper/main.tex
Update paper and supplementary, add main .gitignore

\title[]{Bayesian regression modelling facilitates quantitative
modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

The Laplace approximation yielded XXX samples in XXX time. The
diagnostics indicated that our algorithm was able to find the maximum a
posteriori parameter configuration, approximate the Hessian and use
these quantities to generate approximate posterior samples. The results
can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

Figure~\ref{fig-laplace} summarises the results of comparing the Laplace
approximation of our case study's posterior distribution with the true
posterior. As can be seen from the top left plot, the Laplace method
does not provide a good approximation to the true posterior
distribution, as the marginal distribution of the total log probability
density is clearly different. This was confirmed using the
Kolmogorov-Smirnov test, which is a test to differentiate two empirical
univariate distributions. The two distributions were significantly
different with a p-value indistinguishable from zero.

The difference between the Laplace approximation output and the true
posterior distribution manifests itself not only in the parameter space,
but also in the measurement space. Figure 5 frame B compares the
5\%-95\% interval for flux measurement log likelihoods in the true
posterior with the Laplace approximation; lower log likelihood values
indicate that the modelled and measured values are further away. The
graph shows that the Laplace approximation yielded significantly worse
predictions than the true posterior, even for the training data.

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants {[}LATEX CONSTANT{]}-
and {[}LATEX CONSTANT{]}: see the bottom right cell of figure 5. This
comparison demonstrates that the Laplace method is not able to capture
the correct relationships between parameters' distributions.

This result shows that MCMC, while slower than Laplace approximation, is
unfortunately preferable for posterior inference in this case. We expect
that this is typical of kinetic models of realistic metabolic networks
in general, so we recommend that Maud users use MCMC sampling if
possible.

>>>>

\title[]{Bayesian regression modelling facilitates quantitative
modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

Our overall strategy was to compare samples generated using MCMC, which
we were confident could be treated as draws from the true posterior
distribution, with approximate posterior samples generated using the
Laplace method. We were unable to generate approximate posterior samples
for our main methionine cycle case study using Laplace approximation, as
the algorithm could not recover from solver failures caused by
unrealistic parameter configurations. We therefore made a comparison for
a simpler model. This input can be found at
\url{https://github.com/biosustain/Methionine_model/tree/main/data/example_ode}.

MCMC sampling for our comparison input yielded XXX samples in XXX time;
Laplace sampling yielded XXX samples in XXX time. The diagnostics
indicated that our algorithm was able to find the maximum a posteriori
parameter configuration, approximate the Hessian and use these
quantities to generate approximate posterior samples. The results can be
found at
\url{https://github.com/biosustain/Methionine_model/tree/main/results}.

Figure~\ref{fig-laplace} summarises the results of comparing the samples
generated using each method. As can be seen from the top left plot, the
Laplace method does not provide a good approximation to the true
posterior distribution in this case, as the marginal distribution of the
total log probability density is clearly different. This impression was
confirmed using the Kolmogorov-Smirnov test, which is a test to
differentiate two empirical univariate distributions. The two
distributions were significantly different with a p-value
indistinguishable from zero.

The difference between the Laplace approximation output and the true
posterior distribution manifests not only in the parameter space, but
also in the measurement space. Figure Figure~\ref{fig-laplace} frame B
compares the 5\%-95\% interval for flux measurement log likelihoods in
the true posterior with the Laplace approximation; lower log likelihood
values indicate that the modelled and measured values are further away.
The graph shows that the Laplace approximation yielded significantly
worse predictions than the true posterior, even for the training data.

To further explore why this is the case we compared samples from the
true posterior and the Laplace approximation for the pairwise marginal
distributions of two Michaelis-Menten constants {[}XXX{]}- and
{[}XXX{]}: see the bottom right cell of figure Figure~\ref{fig-laplace}.
This comparison demonstrates that the Laplace method is not able to
capture the correct relationships between parameters' distributions.

This result shows that MCMC, while slower than Laplace approximation, is
unfortunately preferable for posterior inference in this case. We expect
that the Laplace method will produce worse approximations the more
complex the target model. Since the approximation is already
unacceptable for our simple test model, we recommend that Maud users use
MCMC sampling in preference to Laplace approximation if possible when
fitting realistic Bayesian kinetic models.
